---
title: "Usage Guide: Extract, Transform, Load"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage Guide: Extract, Transform, Load}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`epebdmc` makes it easy to work with Brazil's greenhouse gas emissions
data from the First Biennial Transparency Report (BTR1), published by
the Ministry of Science, Technology and Innovation (MCTI/SIRENE).

The package handles the full pipeline: downloading the raw data,
transforming it into a queryable format, and providing simple functions
to filter, visualize, and export results.

## Quick Start
```{r quickstart}
library(epebdmc)

# Step 1: Set up the database (run once)
setup_btr1()

# Step 2: Query, plot, export
get_emissions(sector = "Energy", uf = "SP")
plot_emissions(sector = "Energy", region = "Sudeste")
export_emissions(path = "energy_data.csv", sector = "Energy")
```

## Setup

`setup_btr1()` downloads all 6 sector files from MCTI, validates the
data, and loads everything into a local SQLite database. You only need
to run this once. Run it again with `overwrite = TRUE` when MCTI
publishes updated data.
```{r setup-options}
# Default: downloads from MCTI
setup_btr1()

# Re-run with fresh data
setup_btr1(overwrite = TRUE)

# Use local files instead
setup_btr1(sources = c(
  Energia      = "data/BTR1_UF_Energia.xlsx",
  IPPU         = "data/BTR1_UF_IPPU.xlsx",
  Agropecuaria = "data/BTR1_UF_Agropecuaria.xlsx",
  LULUCF       = "data/BTR1_UF_LULUCF.xlsx",
  Residuos     = "data/BTR1_UF_Residuos.xlsx",
  Total        = "data/BTR1_UF_Total_Brasil.xlsx"
))

# Check what's in the database
get_data_summary()
```

## Querying Data

`get_emissions()` is the main function. All parameters are optional
and can be combined freely.

### Filter by state
```{r query-uf}
# By state code
get_emissions(uf = "RJ")

# By full name
get_emissions(uf = "Rio de Janeiro")

# By IBGE code
get_emissions(uf = 33)

# Multiple states
get_emissions(uf = c("SP", "RJ", "MG"))
```

### Filter by sector
```{r query-sector}
# English or Portuguese names work
get_emissions(sector = "Energy")
get_emissions(sector = "Energia")

# Available sectors: Energy, IPPU, Agriculture, LULUCF, Waste, Total
```

### Filter by gas
```{r query-gas}
# Default is CO2 eq (all gases combined)
get_emissions(uf = "SP")

# Specific gas
get_emissions(uf = "SP", gas = "CH4")

# Multiple gases
get_emissions(uf = "SP", gas = c("CO2", "CH4", "N2O"))

# Available: CO2 eq, CO2, CH4, N2O, HFCs, PFCs, SF6
```

### Filter by year and region
```{r query-year-region}
# Single year
get_emissions(sector = "Energy", year = 2022)

# Year range
get_emissions(sector = "Energy", year = 2010:2022)

# Specific years
get_emissions(sector = "Energy", year = c(1990, 2000, 2010, 2022))

# By region
get_emissions(region = "Nordeste")

# Available regions: Norte, Nordeste, Sudeste, Sul, Centro-Oeste
```

### Include Brasil total
```{r query-brasil}
# By default, the Brasil aggregation row is excluded
# Include it for comparison:
get_emissions(sector = "Energy", year = 2022, includeBrasil = TRUE)
```

### Combine everything
```{r query-combined}
get_emissions(
  sector = "LULUCF",
  region = c("Norte", "Centro-Oeste"),
  gas    = "CO2",
  year   = 2000:2022
)
```

## Visualization

### Time series
```{r plot-line}
# Emissions over time by state
plot_emissions(sector = "Energy", region = "Sudeste")

# Compare sectors for one state
plot_emissions(uf = "SP", colorBy = "sector")

# Compare gases
plot_emissions(
  uf = "SP", sector = "Energy",
  gas = c("CO2", "CH4", "N2O"),
  colorBy = "gas"
)

# By region with total line
plot_emissions(sector = "Energy", colorBy = "region", showTotal = TRUE)
```

### Rankings
```{r plot-ranking}
# Top 10 emitting states
plot_ranking(sector = "Energy", year = 2022, topN = 10)

# Compare sectors
plot_ranking(year = 2022, rankBy = "sector")

# Compare regions
plot_ranking(sector = "LULUCF", year = 2022, rankBy = "region")
```

### Maps

Requires the `geobr` and `sf` packages.
```{r plot-map}
# Choropleth: emissions by state
plot_map(sector = "Energy", year = 2022)

# With values on the map
plot_map(sector = "LULUCF", year = 2022, showValues = TRUE)

# Custom palette
plot_map(sector = "Energy", year = 2022, palette = "Blues")

# Change map: compare two years
plot_map_change(sector = "Energy", yearFrom = 1990, yearTo = 2022)

# LULUCF change with percentages
plot_map_change(
  sector = "LULUCF",
  yearFrom = 2000, yearTo = 2022,
  showValues = TRUE
)
```

## Exporting Data
```{r export}
# CSV
export_emissions(path = "energy_sudeste.csv", sector = "Energy", region = "Sudeste")

# Excel
export_emissions(path = "all_data.xlsx")

# Filtered export
export_emissions(
  path   = "rj_lulucf_2010_2022.csv",
  uf     = "RJ",
  sector = "LULUCF",
  year   = 2010:2022
)
```

## Advanced: Custom Queries

For analyses beyond what the high-level functions provide, use
`get_db()` to access the database directly with `dplyr`.
```{r advanced}
con <- get_db()

# Total emissions by region and sector in 2022
dplyr::tbl(con, "factEmissions") |>
  dplyr::inner_join(dplyr::tbl(con, "dimUf"), by = "ufId") |>
  dplyr::inner_join(dplyr::tbl(con, "dimSector"), by = "sectorId") |>
  dplyr::filter(gasId == 1, yearId == 2022, isBrasilTotal == 0) |>
  dplyr::group_by(regionName, sectorName) |>
  dplyr::summarise(totalKt = sum(emissionKt, na.rm = TRUE)) |>
  dplyr::collect()

db_disconnect(con)
```

## Data Source

All data comes from MCTI's SIRENE platform — the official Brazilian
greenhouse gas inventory system. The BTR1 covers emissions from 1990
to 2022 across the five IPCC sectors (Energy, IPPU, Agriculture,
LULUCF, Waste) for all 27 Brazilian states plus Distrito Federal.

For more information:
[MCTI Climate Page](https://www.gov.br/mcti/pt-br/acompanhe-o-mcti/cgcl/clima)
